Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-21 probabilisticProgrammingMartingales.tex
ongoing
Add ranking function theorems for the confluent semantics.

\begin{theorem} \label{partial implies rankable}
Every partial ranking function is a restriction of a ranking function.
\end{theorem}
\begin{proof}
Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
\begin{align*}
    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
\end{align*}

@@

\begin{theorem}[Antitone ranking function soundness]
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
\end{proof}

@@

\begin{theorem}
  Every antitone partial ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

@@

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{theorem}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{theorem}
\akr{TODO: Move proof of this theorem here, and edit it so that it's actually proving this, rather than trying to prove something stronger.}
\input{y-stopping-time}

@@

@@

\paragraph{}
\akr{For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
\begin{itemize}
    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ %\lo{Define $\sim_N^*$.}
    \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
    \item If $M \to N$ at $\alpha$, no position in any term in $Rch(N)$ is related to $(M,\alpha)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
    \item The subprobability distribution of values obtained by reducing a term in the usual call-by-value semantics is equal to that obtained by any reduction strategy on $\Rightarrow$ (i.e.~a sub-relation of $\Rightarrow$ that is a function) (with some allowance made for the fact that some values are not normal forms of $\Rightarrow$, so the reduction strategy might overshoot the value that $\red^\infty$ reaches. I'm not quite sure what form this will take.)
    \item Similarly, the probability of termination is equal in $\red^\infty$ and any reduction strategy on $\Rightarrow$, so almost sure termination can be equivalently defined based on $\Rightarrow$.
    \item For any closed term $N$, if there is any reduction strategy for $N$ and ranking function corresponding to that reduction strategy, $N$ is AST.
    \item Corresponding sparse and antitone-strict ranking function theorems.
\end{itemize}}

@@

\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}

>>>>

\begin{theorem} \label{thm:partial implies rankable}
Every partial ranking function is a restriction of a ranking function.
\end{theorem}
\begin{proof}
Take a closed term $M$ and partial ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
\begin{align*}
    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
\end{align*}

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
\end{proof}

@@

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone partial ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

@@

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{theorem}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{theorem}

@@

\input{y-stopping-time}

@@

@@

All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too. The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies. They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy). For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x. O]$, and $r(N)$ being undefined for $N$ being a value.

The domain of definition of the ranking functions also needs to be changed from the reachable terms $Rch(M)$ to the \emph{reachable terms with respect to the reduction strategy $r$},
\[
Rch_r(M) = \{N \mid \exists n, N_0, \dots, N_n: N_0 = M, N_n = N, N_i \to N_{i+1} \text{ at } r(N_i)\}.
\]

To be more explicit, the modified forms of the theorems are:
\begin{definition}\rm
A \emph{ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
\begin{itemize}
    \item $f(N) \geq 1+ f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$

    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
\end{itemize}
Any closed term for which a ranking function with respect to a reduction strategy $r$ exists is called \emph{rankable} with respect to $r$. 
\end{definition}

\begin{definition}
A \emph{partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}

\begin{definition}\rm
An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{itemize}
    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$

    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
\end{itemize}
Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
\end{definition}

\begin{definition}
An \emph{antitone partial ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}

\begin{theorem}
If a closed SPCF term $M$ has a ranking function, partial ranking function, antitone ranking function or antitone partial ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
\end{theorem}
\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}

### 2020-12-21 relativised-completeness.tex
Thots on completeness proof: an alternative to ultrafilter on natural numbers.
Add ranking function theorems for the confluent semantics.

\section{Relativised completeness}
\label{sec:Relativised completeness}

If every term reachable from a closed SPCF program is AST then the program is also AST, but not conversely 
(because a diverging term may be reachable with probability 0 from an AST term).

We aim to prove the following conjecture:

\begin{quote}
\emph{Given a closed SPCF term $M$, if every term reachable from $M$ is AST then $M$ admits an antitone ranking function.}
\end{quote}

\begin{enumerate}

\item[(1)] Given an a.s.~finite stopping time $T$ (meaning: $T < \infty$ a.s.) adapted to a filtration, there exist supermartingale $(Y_n)_{n \in \omega}$ adapted to the same filtration and antitone function $\epsilon$ such that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale w.r.t.~$(\epsilon, T)$.

\item[(2)] Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale w.r.t.~$(\epsilon, T)$. 
Then $M$ admits an antitone ranking function.
\end{enumerate}

\lo{Is (1) too general for the conjecture?
Claim (1) assumes an arbitrary a.s.~finite stopping time $T$. This seems too general for the conjecture, which assumes a ``strongly'' AST $M \in \Lambda^0$.
Regardless of the validity of (1), in order to prove the conjecture, I think we would need a more specific version of (1) that assumes a stopping time about terminating runs of $M$ --- such as the stopping time of definition (\ref{eq:y stopping time}) which only counts $\tY$-reduction steps --- so that the antitone-strict supermartingale that is constructed has the necessary structure to yield a ranking function for $M$.}

\iffalse
For (2), note that for every $N \in \mathit{Rch}(M)$, there exists $n \geq 0$ and $s \in S$ such that $\red^n(M, s) = (N, -)$.
Define a function $f : \mathit{Rch}(M) \to \nnReal$ by $f(N) := Y_n'(s)$.
\fi

For 1, consider the following:
\begin{proof}
If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
Otherwise, let $(T_n)_{n \in \omega}$ be defined recursively such that $T_0 = 0$, $T_{n+1} > T_n$ and $\mathbb P[T < T_{n+1}] \leq \frac 1 4 \mathbb P[T < T_n]$. \lo{Since $T$ is a stopping time, we can assume that $T_n \in \mathbb N$. 
Problem with $T_1$: we require $T_1 > 0$ which may contradict the requirement $\mathbb P[T < T_{1}] \leq \frac 1 4 \mathbb P[T < 0] = 0$ for certain $T$, such as those that satisfy $\mathbb P[T < 1] > 0$.
We could define $T_1$ as the least number satisfying $\mathbb P[T < T_1] \geq \frac{1}{4}$.}
We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(Y'_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$ \changed[lo]{a.s.}, $Y_n = Y'_n$ iff $T \geq n$ \changed[lo]{a.s.}, and $Y'_n \geq 2^k$ for $n \geq T_k$.

\lo{$(T_n)_{n \in \omega}$ and $(Y'_n)_{n \in \omega}$ are $\omega$-sequences of numbers. 
There is a convention that random variables are denoted by uppercase letters. 
I suggest we write them as $(t_n)_{n \in \omega}$ and $(y'_n)_{n \in \omega}$ respectively:
in this way, we emphasise that they are sequences of nunbers, and avoid any possible confusion with r.v.}

The antitone function \lo{It is good practice not to begin a sentence with a maths expression / symbol.} $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
\begin{align*}
    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \max \left (\epsilon(2^{k-1}), \frac{2^k}{T_{k+1}-T_k} \right ).
\end{align*}
This ensures that
\begin{itemize}
\item $\epsilon$ is weakly decreasing. 
\item $\epsilon$ is bounded below by 0.
\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{T_{k+1}-T_k}$. 
\lo{This would imply $\frac{2^k}{T_{k+1}-T_k} \geq \epsilon(2^{k-1})$. Why?}
\end{itemize}

$Y'_n$ is then defined recursively by
\begin{align*}
    Y'_0 = 2, \quad
    Y'_{n+1} = (Y'_n - \epsilon(Y'_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
\end{align*}

The fact $Y'_{T_k} \geq 2^{k+1}$ is proven by induction on $k$. For base case, $2 \geq 2$, as required. For the inductive case $k+1$, we first do another induction to prove that $Y'_n \geq 2^k$ for all $T_k \leq n \leq T_{k+1}$. The base case of this inner induction follows from the outer induction hypothesis a fortiori. take the greatest $k$ such that $n > T_k$. By the induction hypothesis, $Y'_{T_k} \geq 2^k$.
\begin{align*}
    Y'_n & = Y'_{T_k} \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} - \sum_{m=T_k}^{n-1} \epsilon(Y'_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \sum_{m=T_k}^{n-1} \epsilon(Y'_m)) \\
    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \sum_{m=T_k}^{n-1} \frac{2^k}{T_{k+1}-T_k}) \\
    & = \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - \frac{2^k (n - T_k)}{T_{k+1}-T_k}) \\
    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (Y'_{T_k} - 2^k) \\
    & \geq \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
    & = \frac{\mathbb P[T > T_k]}{\mathbb P[T > n]} 2^k \\
    & \geq 2^k
\end{align*}
as required. Substituting $n = T_{k+1}$ and using the fact that $\frac{\mathbb P[T > T_k]}{\mathbb P[T > T_{k+1}]}$, the same reasoning gives $Y'_{T_{k+1}} \geq 2^{k+2}$, as required.

Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $Y'_n > 0$ for all n, which implies that $Y_n \geq 0$. The other condition for $(Y_n)$ to be an antitone-strict supermartingale is \akr{The inference from the third to the fourth line is actually incorrect if $\mathcal F$ is not coarsest filtration to which $T$ is adapted. This error may require a substantial change to the proof to correct, but I thought you'd want to see it anyway rather than waiting for me to fix it.} \\
\begin{align*}
    & \expect{Y_{n+1} \mid \mathcal F_n} \\
    = & \expect{Y'_{n+1} 1_{T \geq n+1} \mid \mathcal F_n} \\
    = & Y'_{n+1} \expect{1_{T \geq n+1} \mid \mathcal F_n} \\ 
    = & Y'_{n+1} 1_{T \geq n} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
    = & (Y'_n - \epsilon(Y'_n)) 1_{T \geq n} \\
    = & Y_n - \epsilon(Y'_n) 1_{T \geq n}
\end{align*}
therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
\end{proof}
\lo{Ongoing.}

>>>>

\begin{proof}
If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
Otherwise, let $(t_n)_{n \in \omega}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
We will then define a supermartingale $(Y_n)_{n \in \omega}$ and nonrandom sequence $(y_n)_{n \in \omega}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.

The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
\begin{align*}
    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
\end{align*}
This ensures that
\begin{itemize}
\item $\epsilon$ is weakly decreasing. 
\item $\epsilon$ is bounded below by 0.
\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
\lo{This would imply $\frac{2^k}{t_{k+1}-t_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
\end{itemize}

$y_n$ is then defined recursively by
\begin{align*}
    y_0 = 2, \quad
    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
\end{align*}

The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$. For base case, $2 \geq 2$, as required. For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$. The base case of this inner induction follows from the outer induction hypothesis a fortiori. Take the greatest $k$ such that $n > t_k$. By the induction hypothesis, $y_{t_k} \geq 2^k$.
\begin{align*}
    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
    & \geq 2^k
\end{align*}
as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.

Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all n, which implies that $Y_n \geq 0$. The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
\begin{align*}
    & \expect{Y_{n+1} \mid \mathcal F_n} \\
    = & \expect{y_{n+1} 1_{T \geq n+1} \mid \mathcal F_n} \\
    = & y_{n+1} \expect{1_{T \geq n+1} \mid \mathcal F_n} \\ 
    = & y_{n+1} 1_{T \geq n} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
    = & (y_n - \epsilon(y_n)) 1_{T \geq n} \\
    = & Y_n - \epsilon(y_n) 1_{T \geq n}
\end{align*}
therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
\end{proof}

The assumption that $(\mathcal F_n)_{n \in \omega}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
